\name{}: machine learning over interaction networks and ontology-based background knowledge predicts drug--target interactions

### 2021-04-26 paper/main.tex

We utilize the PhenomeNET ontology \citep{PhenomeNET2011}, an ontology
integrating ontologies such as the Human Phenotype Ontology
\citep{HPO2018}, Gene Ontology \cite{GOoriginal2000, GOrecent2020},
Mammalian Phenotype Ontology \citep{MP2009} and several others.  We
obtained side effects and their links to drugs from SIDER
\citep{SIDER}; SIDER contains side effects encoded using identifiers
from the MedDRA database \citep{MedDRA}. We mapped side effects to the
PhenomeNET ontology using the \textit{Phenomebrowser.net}, which
provides a SPARQL query endpoint for the mentioned resources.

@@

\subsubsection{Combined prediction model}
Combining half-twin and graph convolutional neural networks, we map
all protein representations to their respective node features,
initializing the graph convolutional update steps. The resulting
representations are used for a similarity prediction.
% as presented in
% figure ~\ref{fig:HalfTwinNetwork}
When combining ontology and molecular features with or without the
graph model, we concatenate both protein features and both drugs
features, before plugging them into the graph model for the similarity
computation. An overview of the model architecture, combining both
feature types, is shown in Supplementary Figure
4. Here
the original representations are transformed by a dense layer and then
used as input of a stack (with height 3) of residual graph
convolutional blocks.

@@

\begin{table}[ht]
  \centering
  \begin{tabular}{|p{1.1cm}|p{0.6cm}|p{0.7cm}|p{0.6cm}|p{0.7cm}|p{0.6cm}|p{0.7cm}|p{0.6cm}|p{0.7cm}|}
    \hline
    &\multicolumn{4}{c|}{(a) STITCH results}& \multicolumn{4}{c|}{(b) Yamanishi results}\\
    \hline
    \name{} results&\multicolumn{4}{c|}{PPI graph}&\multicolumn{4}{c|}{PPI graph}\\
    &\multicolumn{2}{c|}{without}&\multicolumn{2}{c|}{with}&\multicolumn{2}{c|}{without}&\multicolumn{2}{c|}{with}\\
    &Macro AUC&Micro $AUC_p$&Macro AUC&Micro $AUC_p$&Macro AUC&Micro AUC&Macro AUC&Micro $AUC_p$\\
    \hline
    MolPred&$0.69$&$0.65$&$0.69$&$0.67$&$0.66$&$0.67$&$0.66$&$0.64$\\
    \hline
    OntoPred&$0.88$&$0.87$&$0.92$&$0.93$&$0.80$&$0.79$&$0.83$&$0.82$\\
    \hline
    \name{} & $0.89$ & $0.90$&$\mathbf{0.93}$&$\mathbf{0.94}$& $0.83$ & $0.82$&$\mathbf{0.84}$&$\mathbf{0.84}$\\
    \hline
  \end{tabular}
  \caption{\label{tab:Results}Results for \name{} on the STITCH
    and Yamanishi datasets evaluated with 5-fold cross-validation. We 
    call the model using only molecular features
    \textit{MolPred} and the model using only ontology-based
    features \textit{OntoPred}. \name{} combines both types of
    features.}
\end{table}

@@

As the difference in performance with different splitting schemes is
quite large, we further evaluated additional drug--target interaction
and drug--target affinity prediction methods that were trained and
evaluation on other datasets.  Following the results of MolTrans
\cite{MolTrans2020}, we reevaluated DeepDTI \cite{DeepDTI2017},
DeepDTA \cite{DeepDTA2018}, DeepConv-DTI \cite{DeepConvDTI2019}, and
MolTrans itself on the BioSnap dataset \cite{BioSnap2018} and compared it to our
``na\"ive'' predictor as well as \name{} (see Table
\ref{tab:comparison2}).  MolTrans was evaluated over the drug--target
pair and the protein split; we were able to reproduce the MolTrans
results (Table \ref{tab:comparison2}), showing a substantial
difference based on the splitting scheme. We additionally computed the
$MicroAUC_p$ score for all considered methods, leading to similar 
results as observed on the Yaminishi dataset. As \name{} considers
less drugs than alternative approaches, we execute a one-tailed T-test
with $\alpha=0.01$ over all considered methods. Considering Macro
AUCs, \name{} improves significantly ($p<0.0001$, one-sided t-test)
over MolTrans, DeepDTI, DeepDTA, and DeepConv-DTI on the BioSnap
dataset, and significantly ($p<0.0001$, one-sided t-test) over DTINet
and DTIGEMS+ on the Yamanishi dataset' DTI-CDF performs
significantly better than \name{} ($p=0.02$, one-sided t-test). % As
                                % we compare \name{} to 4 different
                                % methods over BioSnap and 3 over
                                % Yamanishi dataset, we add a
                                % Bonferroni correction by multiplying
                                % the $p$-values by 4 and 3,
                                % respectively, leading to $p<0.0002$
                                % ($\alpha=0.01$) for all previous
                                % comparisons except DTI-CDF.
We further run another one-tailed t-test concerning $MicroAUC_p$ with $\alpha=0.01$ including Bonferroni correction. We find that \name{} advances significantly over MolTrans ($p<0.00001$), DeepDTI ($p<0.00001$), DeepDTA ($p<0.00001$), and DeepConv-DTI ($p<0.00001$) regarding BioSnap dataset, and over DTINet ($p<0.00001$), DTIGEMS+ ($p<0.00001$) and DTI-CDF ($p<0.00001$) regarding Yamanishi dataset.

>>>>

We utilize the PhenomeNET ontology \citep{PhenomeNET2011}, an ontology
integrating ontologies such as the Human Phenotype Ontology
\citep{HPO2018}, Gene Ontology \cite{GOoriginal2000, GOrecent2020},
Mammalian Phenotype Ontology \citep{MP2009} and several others.  We
obtained side effects and their links to drugs from SIDER
\citep{SIDER}; SIDER contains side effects encoded using identifiers
from the MedDRA database \citep{MedDRA}. We mapped side effects to the
PhenomeNET ontology using the \textit{Phenomebrowser.net}, which
provides a SPARQL query endpoint for the mentioned resources. The overall structure is depicted in Supplementary Figure 1.

@@

\subsubsection{Combined prediction model}
Combining half-twin and graph convolutional neural networks, we map
all protein representations to their respective node features,
initializing the graph convolutional update steps. The resulting
representations are used for a similarity prediction.
% as presented in
% figure ~\ref{fig:HalfTwinNetwork}
When combining ontology and molecular features with or without the
graph model, we concatenate both protein features and both drugs
features, before plugging them into the graph model for the similarity
computation. An overview of the model architecture, combining both
feature types, is shown in Supplementary Figure
4. Here
the original representations are transformed by a dense layer and then
used as input of a stack (with height 3) of residual graph
convolutional blocks. The full eventual model of \name{} is shown in Supplementary Figure 4.

@@

\begin{table}[ht]
  \centering
  \begin{tabular}{|p{1.1cm}|p{0.6cm}|p{0.7cm}|p{0.6cm}|p{0.7cm}|p{0.6cm}|p{0.7cm}|p{0.6cm}|p{0.7cm}|}
    \hline
    &\multicolumn{4}{c|}{(a) STITCH results}& \multicolumn{4}{c|}{(b) Yamanishi results}\\
    \hline
    \name{} results&\multicolumn{4}{c|}{PPI graph}&\multicolumn{4}{c|}{PPI graph}\\
    &\multicolumn{2}{c|}{without}&\multicolumn{2}{c|}{with}&\multicolumn{2}{c|}{without}&\multicolumn{2}{c|}{with}\\
    &Macro AUC&Micro $AUC_p$&Macro AUC&Micro $AUC_p$&Macro AUC&Micro AUC&Macro AUC&Micro $AUC_p$\\
    \hline
    MolPred&$0.69$&$0.65$&$0.69$&$0.67$&$0.66$&$0.67$&$0.66$&$0.64$\\
    \hline
    OntoPred&$0.88$&$0.87$&$0.92$&$0.93$&$0.80$&$0.79$&$0.83$&$0.82$\\
    \hline
    \name{} & $0.89$ & $0.90$&$\mathbf{0.93}$&$\mathbf{0.94}$& $0.83$ & $0.82$&$\mathbf{0.84}$&$\mathbf{0.84}$\\
    \hline
  \end{tabular}
  \caption{\label{tab:Results}Results for \name{} on the STITCH
    and Yamanishi datasets evaluated with 5-fold cross-validation. We 
    call the model using only molecular features
    \textit{MolPred} and the model using only ontology-based
    features \textit{OntoPred}. \name{} combines both types of
    features. For a comparison of MicroAUC$_p$ and MicroAUC$'_p$ (See Section \ref{sec:evaluation} for more information) see Supplementary Table 2. }
\end{table}

@@

As the difference in performance with different splitting schemes is
quite large, we further evaluated additional drug--target interaction
and drug--target affinity prediction methods that were trained and
evaluated on other datasets. Following the results of MolTrans
\cite{MolTrans2020}, we reevaluated DeepDTI \cite{DeepDTI2017},
DeepDTA \cite{DeepDTA2018}, DeepConv-DTI \cite{DeepConvDTI2019}, and
MolTrans itself on the BioSnap dataset \cite{BioSnap2018} and compared it to our
``na\"ive'' predictor as well as \name{} (see Table
\ref{tab:comparison2}).  MolTrans was evaluated over the drug--target
pair and the protein split; we were able to reproduce the MolTrans
results (Table \ref{tab:comparison2}), showing a substantial
difference based on the splitting scheme. We additionally computed the
$MicroAUC_p$ score for all considered methods, leading to similar 
results as observed on the Yaminishi dataset. As \name{} considers
less drugs than alternative approaches, we execute a one-tailed T-test
with $\alpha=0.01$ over all considered methods. Considering Macro
AUCs, \name{} improves significantly ($p<10^{-4}$, one-sided t-test)
over MolTrans ($t=36.6$), DeepDTI ($t=40.5$), DeepDTA ($t=36.6$), and DeepConv-DTI ($t=40.5$) on the BioSnap
dataset, and significantly ($p<10^{-4}$, one-sided t-test) over DTINet ($t=20.8$)
and DTIGEMS+ ($t=18.0$) on the Yamanishi dataset' DTI-CDF performs
significantly better than \name{} ($p=0.02$, one-sided t-test). % As
                                % we compare \name{} to 4 different
                                % methods over BioSnap and 3 over
                                % Yamanishi dataset, we add a
                                % Bonferroni correction by multiplying
                                % the $p$-values by 4 and 3,
                                % respectively, leading to $p<0.0002$
                                % ($\alpha=0.01$) for all previous
                                % comparisons except DTI-CDF.
We further run another one-tailed t-test concerning $MicroAUC_p$ with $\alpha=0.01$ including Bonferroni correction. We find that \name{} advances significantly over MolTrans ($p<10^{-4}, t=21.2$), DeepDTI ($p<10^{-4}, t=31.0$), DeepDTA ($p<10^{-4}, t=33.4$), and DeepConv-DTI ($p<10^{-4}, t=23.7$) regarding BioSnap dataset, and over DTINet ($p<10^{-4}, t=29.6$), DTIGEMS+ ($p<10^{-4}, t=28.0$) and DTI-CDF ($p<10^{-4}, t=9.5$) regarding Yamanishi dataset.
